Investigating magnetic dynamo evolution with \TESS\ field dwarfs

### 2017-10-06 ms.tex
talk about vetting

% Still to do: estimate the number of 16th mag stars and brighter in the TESS
% southern sky. ~ 15 million, call it 10.
% Figure out how to incorporate the target list.
% Get detection probabilities for stars on the list
% Apply Derek's corrections
% Calculate the distance at which you get a ~ 15th mag at ~1kpc
% cite the sullivan paper.
% Mention vetting.

@@

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport found this period bimodality extends throughout all
masses in the \Kepler\ rotation sample for nearby stars \citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The \TESS\ FFI targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a discontinuous age distribution,
the feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the \TESS\ FFI targets to the upcoming second data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
For the all-southern-hemisphere  analysis we will focus on F and G stars since
the 27.4 day baseline of \TESS\ fields will limit measurable rotation periods
to less than around 15 days.
The longer time coverage in the CVZ will allow us to study the rotation
period bimodality in K and M dwarfs, since the bimodality appears at longer
rotation periods for these lower mass stars.
Unlike \kepler\ and \ktwo\ light curves, {\bf \TESS\ provides the all-sky
coverage necessary to map the period bimodality across the galaxy.}

@@

Although the classical spin-down law of \citet{skumanich1972} (Period
$\propto$ Age$^{1/2}$) holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods, with the maximum
measurable period strongly depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the Continuous Viewing Zones (CVZs).
Many of these will have spectra from the \TESS --\HERMES\ survey, from which
it will be possible to calculate isochronal ages, enabling further calibration
of the gyrochronology relations.
{\bf We propose to calibrate gyrochronology at old ages for FGKM stars using
the isochronal ages of \TESS--\hermes\ stars.}
We will also use the rotation periods of \TESS\ field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many stars observable by \tess\ will have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action-angle coordinates of the stars, both of which are age
indicators.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
{\bf We propose to calibrate the gyrochronology relations at old ages for FGKM
stars using vertical actions of \Gaia\ DR2 targets in the CVZ.}

@@

The full set of available \TESS\ FFI light curves will be exceedingly large
and we intend to prioritize certain groups of stars for period analysis,
calculating a simple (and computationally fast) Lomb-Scargle periodogram for
the low priority objects and running a full probabilistic, Gaussian process
period analysis on the high priority targets.
High priority targets include: comoving pairs, stars in the CVZs (especially
those with spectra), stars in TGAS, stars at large distances, plus any
potential asteroseismic dwarfs and exoplanet hosts.
Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
expect to be able to measure rotation periods for FFI stars down to 16th
magnitude.
We note, however that not all stars will have measurable rotation periods,
even if they are bright enough --- not all stars show rotational variability
in their light curves either because they are pole-on or because they are
inactive.

>>>>

% Still to do: estimate the number of 16th mag stars and brighter in the TESS
% southern sky. ~ 15 million, call it 10.
% Cite the sullivan paper.
% Figure out how to incorporate the target list.
% Get detection probabilities for stars on the list
% Calculate the distance at which you get a ~ 15th mag at ~1kpc
% Mention vetting.

@@

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport found this period bimodality extends throughout all
masses in the \Kepler\ rotation sample for nearby stars \citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The \TESS\ FFI targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a discontinuous age distribution,
the feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the \TESS\ FFI targets to the upcoming second data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
For the all-southern-hemisphere  analysis we will focus on F and G stars since
the 27.4 day baseline of \TESS\ fields will limit measurable rotation periods
to less than around 15 days.
The longer time coverage in the Continuous Viewing Zone (CVZ) will allow us to
study the rotation period bimodality in K and M dwarfs, since the bimodality
appears at longer rotation periods for these lower mass stars.
Unlike \kepler\ and \ktwo\ light curves, {\bf \TESS\ provides the all-sky
coverage necessary to map the period bimodality across the galaxy.}

@@

Although the classical spin-down law of \citet{skumanich1972} (Period
$\propto$ Age$^{1/2}$) holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods, with the maximum
measurable period strongly depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the CVZ.
Many of these will have spectra from the \TESS --\HERMES\ survey, from which
it will be possible to calculate isochronal ages, enabling further calibration
of the gyrochronology relations.
{\bf We propose to calibrate gyrochronology at old ages for FGKM stars using
the isochronal ages of \TESS--\hermes\ stars.}
We will also use the rotation periods of \TESS\ field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many stars observable by \tess\ will have proper motions, parallaxes,
positions and radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action-angle coordinates of the stars, both of which are age
indicators.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
{\bf We propose to calibrate the gyrochronology relations at old ages for FGKM
stars using vertical actions of \Gaia\ DR2 targets in the CVZ.}

@@

Based on \citet{sullivan2015}, we expect around 10 million \TESS\ FFI
targets brighter than 15th magnitude in the southern ecliptic hemisphere.
We intend to prioritize certain groups of stars for period analysis,
calculating a simple (and computationally fast) Lomb-Scargle periodogram for
the low priority objects and running a full probabilistic, Gaussian process
period analysis on the high priority targets.
High priority targets will include: comoving pairs, stars in the CVZ
(especially those with spectra), stars in TGAS, stars at large distances, plus
any potential asteroseismic dwarfs and exoplanet hosts.
Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
expect to be able to measure rotation periods for FFI stars down to 15th
magnitude.
We note, however that not all stars will have measurable rotation periods,
even if they are bright enough --- not all stars show rotational variability
in their light curves either because they are pole-on or because they are
inactive.
We will vet incorrect rotation period measurements using a combination of a
Lomb-Scargle periodogram and an autocorrelation function by
training on a set of light curves with a range of variability amplitudes and
periods.
However, based on the high success rates of a Lomb-Scargle periodogram method
with high-pass filtered light curve and the Gaussian process method
demonstrated in \citet{angus2017}, we expect a low rate of incorrect rotation
periods.
